---
title: "Use seurat scaling on ADT"
author: "Stephany Orjuela & Anthony Sonrel"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: show
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "pdf"))
execute <- FALSE
```

## Load packages


```{r}
suppressPackageStartupMessages({
 library(ggplot2)
 library(SingleCellExperiment)
 library(Seurat)
 library(sctransform)
 library(scater)
 })
```

## Import, convert
```{r}
d <- readRDS("sce/d_rna_filt.rds")

altExp(d)
d.seurat <- as.Seurat(d)

d.seurat[["ADT"]] <- CreateAssayObject(counts = counts(altExp(d)))

```


## Normalize, scale

Setting center to TRUE will center the expression for each feautre by subtracting the average expression for that feature.
Setting scale to TRUE will scale the expression level for each feautre by dividing the centered feature expression levels by 
their standard deviations if center is TRUE and by their root mean square otherwise.
```{r}
d.seurat <- NormalizeData(d.seurat, verbose = FALSE, normalization.method = "CLR", 
                          assay = "ADT",
                          margin = 2)
d.seurat <- ScaleData(d.seurat, 
                      #features = rownames(d.seurat@assays$ADT@data), 
                      assay = "ADT", verbose = FALSE,
                      do.scale = TRUE,
                      do.center = TRUE,
                      split.by = "Sample"
                      )

adt.data <- GetAssayData(d.seurat, slot = "scale.data", assay = "ADT")

idx <- match(colnames(adt.data),colnames(d))
rownames(adt.data) <- rownames(altExp(d))
assay(altExp(d), "seurat.scale") <- adt.data[,idx]
```

add that to a SCE for easier viz
```{r, fig.width = 10, fig.height= 10}
library(CATALYST)

#add marker types
rowData(altExp(d))$marker_class <- "type"
rowData(altExp(d))$marker_class[grepl("IgG", rownames(altExp(d)))] <- "none"
colData(altExp(d)) <- colData(d)[,c("sample_id", "seurat", "Sample", "patient_id", "condition","total", "batch.discard",
  grep("altexps_", colnames(colData(d)), value = TRUE))]

rowData(altExp(d))$marker_name <- rowData(altExp(d))$ID

#turn coldata into factors
colData(altExp(d))$sample_id <- factor(colData(altExp(d))$sample_id)
colData(altExp(d))$patient_id <- factor(colData(altExp(d))$patient_id)
colData(altExp(d))$condition <- factor(colData(altExp(d))$condition)
colData(altExp(d))$Sample <- factor(colData(altExp(d))$Sample)

#run UMAP
set.seed(1010010)
altExp(d) <- runDR(altExp(d), cells = 2000, features = "type", assay = "seurat.scale")

#UMAP dimension reduction showing cells per sample ID
plotDR(altExp(d), color_by="sample_id", assay = "seurat.scale")

#UMAP dimension reduction showing cells per well
plotDR(altExp(d), color_by="Sample", assay = "seurat.scale")


#look at all markers
#Heatmap of median scaled expression of the ADT markers
plotExprHeatmap(altExp(d), features = "type", assay = "seurat.scale",
                bars = TRUE, row_clust = TRUE, scale = "never")

#Expression of scaled markers in UMAP plot
plotDR(altExp(d), color_by = type_markers(altExp(d)), assay = "seurat.scale")

assay(altExp(d), "exprs") <- as.matrix(assay(altExp(d), "seurat.scale"))
#Smoothed density of seurat-scaled markers expression per well
plotExprs(altExp(d), color_by = "Sample") + coord_cartesian(xlim =c(-5,5))

# CD45 vs total UMI
total_expr <- colSums(assay(altExp(d), "counts"))
num_features <- colSums(assay(altExp(d), "counts") > 0)
df <- data.frame(total_expr = total_expr,
                 num_features = num_features,
                 #ccr5_clr = assay(altExp(d), "seurat.scale")["CCR5",],
                 ccr5_clr = assay(altExp(d), "seurat.scale")["CD45",],
                 sample_id = colData(altExp(d))$sample_id,
                 Sample = colData(altExp(d))$Sample)
#CD45 scaled expression vs total UMI, colored per well
ggplot(df, aes(log10(total_expr), ccr5_clr, colour=Sample)) +
  geom_point() + xlim(2,4) + facet_wrap(~Sample, nrow=2) +
  geom_density2d(colour="white") +
  geom_vline(aes(xintercept = 2.75, colour="black")) +
  geom_hline(aes(yintercept = 0, colour="black"))


cd45 <- t(assay(altExp(d), "seurat.scale"))[,"CD45"]
others <- t(assay(altExp(d), "seurat.scale"))
others <- others[,-which(colnames(others) == "CD45")]
nam <- rep(colnames(others), each = nrow(others))
others <- as.numeric(others)
df <- data.frame(CD45 = rep(cd45, length(unique(nam))), 
                 markers_expr = others, 
                 markers = nam)
# CD45 normalized expression vs other markers
ggplot(df, aes(markers_expr, CD45)) +
  xlim(min(cd45), max(cd45)) + 
  ylim(min(others), max(others)) + 
  facet_wrap(~markers, nrow=6) +
  geom_bin2d() + xlab("") +
  scale_fill_continuous(type = "viridis")+
  geom_vline(aes(xintercept = 0)) +
  geom_hline(aes(yintercept = 0))



```


## Cluster on scaled counts

We are performing here louvain clustering based on the seurat-scaled expression of the ADT markers.


**The resulting clustering will be used in the downstream analysis (annotation, DS,..)**

```{r}

library(scran)
set.seed(1010001)
g.adt <- buildSNNGraph(altExp(d), d=NA, assay.type = "seurat.scale", k = 8, 
                       subset.row = type_markers(altExp(d)), 
                       BPPARAM = BiocParallel::MulticoreParam(5)) 

clusters.adt <- igraph::cluster_louvain(g.adt)
altExp(d)$cluster_id <- factor(clusters.adt$membership)
altExp(d)$cluster_louvain <- factor(clusters.adt$membership)
table(altExp(d)$cluster_louvain)
```

```{r, fig.width = 10, fig.height= 10}
#change clust labels to plot
source("scripts/getUMAPclusters.R")
by_text <- getUMAPclusters(altExp(d)) 

#UMAP dimension reduction colored by cluster
plotDR(altExp(d), color_by="cluster_id", assay = "seurat.scale") + 
  geom_point(alpha = 0.3, size = 0.5) +
  annotate("text", x=by_text$x, y=by_text$y, label=names(by_text$x))

metadata(altExp(d))$cluster_codes <- data.frame(
  custom = factor(clusters.adt$membership), 
  levels = clusters.adt$membership)

#Abundance plot of clusters, separated by sample
plotAbundances(altExp(d), k = "custom", by = "sample_id", 
               group_by = "condition")

#Expression heatmap of seurat-scaled markers, aggregated *without scaling*
plotExprHeatmap(altExp(d), by = "cluster_id", assay = "seurat.scale",
                k = "custom", features = "type", 
                scale = "never", bars = TRUE, row_clust = FALSE)
#Expression heatmap of seurat-scaled markers, *scaled, trimed and then aggregated per cluster*
plotExprHeatmap(altExp(d), by = "cluster_id", assay = "seurat.scale",
                k = "custom", features = "type", 
                scale = "first", bars = FALSE, row_clust = FALSE)
#Expression heatmap of seurat-scaled markers, *aggregated then scaled and trimmed per cluster*
plotExprHeatmap(altExp(d), by = "cluster_id", assay = "seurat.scale",
                k = "custom", features = "type", 
                scale = "last", bars = FALSE, row_clust = FALSE)
##save as heatmap_celltypes_scale

```



```{r}

saveRDS(d, "sce/d_ADT_seuratscaled.rds")


```



```{r}
sessionInfo()
```

