---
title: "Look at Expression markers in celltypes"
author: "Stephany Orjuela & Anthony Sonrel"
date: "2021-02-17"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: show
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "pdf"))
torun = TRUE
```

## Load packages



```{r}
suppressPackageStartupMessages({
 library(SingleCellExperiment)
 library(scater)
 library(purrr)  
 library(data.table)
 library(ComplexHeatmap)
 library(CATALYST)
 library(Seurat)
 })

```


```{r, fig.width = 10, fig.height= 10}
#load object
annots <-  read.csv("metadata/annotation.csv", header = TRUE, stringsAsFactors = FALSE)

#load RNA-filtered SCE
d <- readRDS("sce/d_adt_clust.rds")
#d <- d[, Cells(umap)]

idx <- match(colnames(d), annots$cell)
colData(d)$predicted.id <- annots$predicted.celltype.l2[idx]
table(d$predicted.id)
```


```{r}
#remove clusters with < 10 cells
d <- d[,!d$predicted.id %in% names(which(table(d$predicted.id)< 10))]

saveRDS(d, "sce/d_adt_annot.rds")
```

## Look at RNA UMAP with annot annotations

```{r, fig.width = 10, fig.height= 10}
source("scripts/getUMAPclusters.R")
by_text <- getUMAPclusters(d, coldata = "predicted.id", meta = NULL) 
plotDR(d, color_by = "predicted.id", assay = "logcounts", dr = "UMAP") +
  geom_point(alpha = 0.3, size = 0.5) +
  annotate("text", x=by_text$x, y=by_text$y, label=names(by_text$x), size = 3)
```


```{r, fig.width = 10, fig.height= 10}
#annot projection
by_text <- getUMAPclusters(d, coldata = "predicted.id", meta = NULL, dr = "UMAP") 
plotDR(d, color_by = "predicted.id", assay = "logcounts", dr = "UMAP") +
  geom_point(alpha = 0.3, size = 0.5) +
  annotate("text", x=by_text$x, y=by_text$y, label=names(by_text$x), size = 3)
```


## Get RNA markers per celltype (DE analysis between celltypes)

There are different ways to do this. Here we test two approaches. Whether the highly ranked genes are those that are DE between the current group and any other group ("any") or all other groups ("all"). 

With `all` it's easier to visualize and more strict, because the null hypothesis is that the gene is not DE in all contrasts (1 cell type Vs each of the other). 

With `any` it does not explicitly favour genes that are uniquely expressed in a cluster. Rather, it focuses on combinations of genes that - together - drive separation of a cluster from the others.

*About the summary.logFC*:
in `all`: is defined as the effect size from the pairwise comparison with the largest p-value. This reflects the fact that, with this approach, a gene is only as significant as its weakest DE.

in `any`: is defined as the effect size from the pairwise comparison with the lowest p-value. The combined p-value is computed by applying Simes' method to all p-values.


```{r}
#what are the highest expressed genes overall
#plotHighestExprs(d, colour_cells_by= "Sample")

# find markers
d <- d[!is.na(rowData(d)$symbol),]
rownames(d) <- rowData(d)$symbol
testf <- "all" #all
markers <- scran::findMarkers(d,
                              pval.type = testf,
                              direction = "up",
                              groups = d$predicted.id, 
                              block = d$Sample)



#organize tables to export in csv per celltype
celltypes <- names(markers)
dir.create("reports/TABLES/")

samecols <- sapply(celltypes, function(u){
    subdf <- data.frame(markers[[u]][,c("p.value","FDR","summary.logFC")])
    
    #choose markers below cutoff
    subdf <- subdf[subdf$FDR < 0.05,]
    
    #only export tables with something
    if(nrow(subdf) > 0) {
    write.csv(subdf, 
              file = sprintf("reports/TABLES/DE_genes_against_%s_%s.csv",
                             testf, u))
    }
  })

#toexport <- dplyr::bind_rows(samecols, .id = "celltype")
#write.csv(toexport, file = sprintf("DE_genes_against_%s_celltype.csv", testf))

#get genes to plot
mark_per_clus <- lapply(markers, function(u){
    genes <- rownames(u[u$FDR < 0.05,])
    if(length(genes) > 0) {
    if(length(genes) >= 10) genes[1:10] else genes
    }
  })

mark_per_clus
```



```{r}
umarks <- unique(unlist(mark_per_clus))
```

### Heatmaps

```{r, fig.width = 10, fig.height= 10}
#add metadata to sce
metadata(d)$cluster_codes <- data.frame(
  custom = factor(1:length(unique(d$predicted.id))),
  annot = factor(unique(d$predicted.id)))

idx <- match(d$predicted.id, metadata(d)$cluster_codes$annot)
d$cluster_id <- metadata(d)$cluster_codes$custom[idx]

#set to plot RNA heatmap
dsub <- d[umarks, ]
colData(dsub) <- colData(dsub)[,c("Sample", "sample_id", "patient_id", "condition", "cluster_id")]

#sample heatmaps
# plotExprHeatmap(dsub, assay = "logcounts", by = "sample_id",
#                 bars = TRUE, row_clust = FALSE, scale = "last")

#cluster heatmap
plotExprHeatmap(dsub, by = "cluster_id", k = "custom", m = "annot",
                scale = "never", row_clust = FALSE, assay = "logcounts")
```


```{r, fig.width = 10, fig.height= 10}
plotExprHeatmap(dsub, by = "cluster_id", k = "custom", m = "annot",
                scale = "first", row_clust = FALSE, assay = "logcounts")
```


Plot too big for RMD report (see `heatmaps` folder)

```{r, eval = FALSE, echo  = FALSE}

#df for rowannotation
dfrow <- data.frame(cluster_id = rep(names(mark_per_clus), 
                                     lengths(mark_per_clus)))

#subset to only clusters with some DE genes
cellstouse <- names(mark_per_clus)[!sapply(mark_per_clus, is.null)]
dsub <- d[,d$predicted.id %in% cellstouse]

suppressPackageStartupMessages({
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
})
dir.create("reports/HEATMAPS")
#png("test.png", width = 1000, height = 1500)
pdf("reports/HEATMAPS/heatmap_markers_long.pdf"), width = 11, height = 14)
source("scripts/plotExprHeatmapLarge.R")
set.seed(111000)
plotExprHeatmapFull(dsub, features = umarks, 
                    drow = dfrow, #add celltypes for gene markers
                    #annot_cols = c("cluster_ADT", "Sample"),
                    annot_cols = "predicted.id",
                    col_clust = FALSE,
                    row_clust = FALSE,
                    assay = "logcounts",
                    cells = 100,
                    k = "predicted.id",
                    row_dend = FALSE, col_dend = FALSE,
                    split = dsub$predicted.id)

#on scaled ADTs
altExp(d)$annot <- factor(cluster_ids(altExp(d), "merging_louvain"))

plotExprHeatmapFull(altExp(d), features = type_markers(altExp(d)), 
                    #annot_cols = c("cluster_ADT", "Sample"),
                    annot_cols = "annot",
                    col_clust = FALSE,
                    assay = "seurat.scale",
                    cells = 100,
                    k = "annot",
                    row_dend = FALSE, col_dend = FALSE,
                    split = altExp(d)$annot)

#on non-scaled normalized ADTs
clr_function <- function(x) {
          return(log1p(x = x / (exp(x = sum(log1p(x = x[x > 0]), na.rm = TRUE) / length(x = x)))))
}
ab_norm <- apply(counts(altExp(d)), 2, clr_function)
assay(altExp(d), "CLR") <- ab_norm
plotExprHeatmapFull(altExp(d), features = type_markers(altExp(d)), 
                    #annot_cols = c("cluster_ADT", "Sample"),
                    annot_cols = "annot",
                    col_clust = TRUE,
                    assay = "CLR",
                    cells = 100,
                    k = "annot",
                    row_dend = FALSE, col_dend = FALSE,
                    split = altExp(d)$annot)
```



```{r}
sessionInfo()
```

