---
title: "Monocle3 pseudotime analysis"
author: "Anthony Sonrel"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: hide
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = c("png", "pdf"), warning=FALSE)
```


```{r, warning=FALSE}
suppressPackageStartupMessages({
 library(SingleCellExperiment)
 library(scater)
 library(purrr)  
 library(data.table)
 library(ComplexHeatmap)
 library(CATALYST)
 library(monocle3)
 library(scran)
 library(patchwork)
 library(ggridges)
 })


#load object after manual annotation
d <- readRDS("sce/d_adt_annot.rds")

```

```{r}
## merge clusters with manual clustering
mergetab <- read.csv("metadata/annotation.csv", header = TRUE, stringsAsFactors = FALSE)
```


```{r}
metadata(altExp(d))$cluster_codes <- data.frame(
  custom = factor(levels(altExp(d)$cluster_id)))

metadata(altExp(d))$cluster_codes$merging_louvain <- factor(mergetab$new_cluster)
```

## MONOCLE 3 pseudotime analysis


The analysis is mostly based on : https://cole-trapnell-lab.github.io/monocle3/docs/. 



```{r}

## rna
rowData(d)$gene_short_name <- paste0(rowData(d)$symbol, ".", rowData(d)$ID)
cds <- new_cell_data_set(counts(d), cell_metadata = colData(d), gene_metadata = rowData(d))
colData(cds)$cluster_id <- colData(altExp(d))$cluster_id
colData(cds)$cluster_name <- plyr::mapvalues(colData(cds)$cluster_id, 
                                       from = as.numeric(as.character(cluster_codes(altExp(d))$custom)), 
                                       to = as.character(cluster_codes(altExp(d))$merging_louvain))
colData(cds)$cluster_id <- droplevels(colData(cds)$cluster_id )
colData(cds)$cluster_name <- droplevels(colData(cds)$cluster_name )

```


`Monocle` pseudotime analysis requires separate preprocessing steps: 

- normalization and dimension reduction

- batch effect removal (using `batchelor`)

- clustering

i.e., **it will not use the preprocessing (normalization, clustering) that was done in our analysis and start again from the raw counts data.** 

## Trajectory graph 

Below, a UMAP plot of the data. It is different from the other UMAP plots of the analysis as we are using `Monocle`-specific functions. The groups of cells that we can observe is also not well overlapping with the cell labels as `Monocle` performed UMAP on the **RNA information**, while the cell labels were inferred from the **protein information**. 

```{r, fig.width=9, fig.height=6}


# Select HVG to speed up MONOCLE process ()
batch <- d$Sample
dec.d_rna <- modelGeneVar(d, block = batch)
chosen.hvgs <- getTopHVGs(dec.d_rna, n=1000)

## Step 1: Normalize and pre-process the data
cds <- preprocess_cds(cds, num_dim = 100, 
                      use_genes = chosen.hvgs)
## Step 2: Remove batch effects with cell alignment
cds <- align_cds(cds, alignment_group = "Sample")
## Step 3: Reduce the dimensions using UMAP
cds <- reduce_dimension(cds, cores = 10)



```

Then, `Monocle` clusters the data and fit a principal graph on each partition of the data, which will be required for the pseudotime analysis:

```{r, fig.width=9, fig.height=6}


## Step 4: Cluster the cells
cds <- cluster_cells(cds, k = 50) # increased k to have lower resolution

## Step 5: Learn a graph
# use_partition = false to connect branches, if needed
cds <- learn_graph(cds) 

plot_cells(cds,
           color_cells_by = "cluster_name",
           label_cell_groups = FALSE,
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE, label_principal_points = TRUE) 


```



```{r, fig.width=9, fig.height=6}
## Step 6: Order cells
# a helper function to identify the root principal points:
ROOT = "XXX"
get_earliest_principal_node <- function(cds, root=ROOT){
  cell_ids <- which(colData(cds)[, "cluster_name"] == root)
  
  closest_vertex <-
  cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <-
  igraph::V(principal_graph(cds)[["UMAP"]])$name[as.numeric(names
  (which.max(table(closest_vertex[cell_ids,]))))]
  
  root_pr_nodes
}
# For RNA, choose anchor points
cds <- order_cells(cds, root_pr_nodes=c("AA", "BB", "CC"))


## plotting and showing pseudotime
# grey cells: lack information to infer pseudotime
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=1.5) 
```

Plotting the Monocle UMAP, colored by cDC type and pseudotime next to each other for better comparison: 

```{r, fig.width=8, fig.height=4}
plot_cells(cds,
           color_cells_by = "cluster_name",
           label_cell_groups = FALSE,
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE) + 
plot_cells(cds,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=1.5) 
```

Plotting the Monocle UMAP for cDC cells, highlitghting the sample of origin and condition. This is used to check if condition or sample drives the clusters that we observe on the UMAP: 

```{r, fig.width=9, fig.height=6}
# plot_cells(cds[ ,sample(1:ncol(cds), ncol(cds))],
#            color_cells_by = "Sample",
#            label_cell_groups = FALSE,
#            label_groups_by_cluster=FALSE,
#            label_leaves=FALSE,
#            label_branch_points=FALSE, 
#            alpha = 0.5) + 
plot_cells(cds[ ,sample(1:ncol(cds), ncol(cds))],
           color_cells_by = "condition",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=1.5, 
           alpha = 0.5) 
```

```{r, fig.width=8, fig.height=6}

plts <- lapply(unique(colData(cds)$condition), function(x){
  
  plot_cells(cds[, colData(cds)$condition == x],
           color_cells_by = "condition",
           label_cell_groups = FALSE,
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE, 
           alpha = 0.5) 
  
})
patchwork::wrap_plots(plts)
```


```{r, include=FALSE, eval=FALSE}
## 3D plots, not relevant ATM
# cds_3d <- reduce_dimension(cds, max_components = 3)
# cds_3d <- cluster_cells(cds_3d, k = 50)
# cds_3d <- learn_graph(cds_3d, use_partition = FALSE)
# cds_3d <- order_cells(cds_3d, root_pr_nodes=get_earliest_principal_node(cds_3d, ROOT))
# colData(cds_3d)$cluster_name <- as.character(colData(cds_3d)$cluster_name)
# cds_3d_plot_obj <- plot_cells_3d(cds_3d, color_cells_by="cluster_name",)
# cds_3d_plot_obj
```

## Population density 

Density of cell-type populations along the pseudotime. The populations in the early pseudotime correspond to the *anchors* that were used to define the graph and should correspond to the early cell populations. 

```{r, fig.width=8, fig.height=8}


psuper <- function(cds, var_plot){
  proj_dt <- data.frame("psuper" = cds@principal_graph_aux[["UMAP"]]$pseudotime,
                        "label_input" = colData(cds)[, var_plot])
  # proj_test <- proj_dt#[-which(proj_dt$psuper == Inf),]
  # new_max <- max(proj_test$psuper) + 2*sd(proj_test$psuper)
  # proj_dt[which(proj_dt$psuper == Inf),"psuper"] <- new_max
  # proj_dt[which(proj_dt$psuper == Inf),"psuper"] <- new_max
  # plot
    ggplot(proj_dt) +
        aes( x=psuper, y=label_input, label_input, fill=label_input, colour=label_input ) +
        geom_density_ridges( alpha=0.5 ) +
        # scale_fill_manual( values=col_group ) +
        #geom_vline( data=cuts_dt, aes(xintercept=psuper, colour=label_input) ) +
        # scale_colour_manual( values=col_group ) +
        guides(
            fill    = guide_legend(override.aes = list(alpha=1), reverse = TRUE)
            ,colour = FALSE
            ) +
        labs(
            x       = 'Pseudotime'
            ,y      = 'Density'
            ,fill   = 'cluster_name'
            ) +
        theme_bw()
}
psuper(cds, "cluster_name") 

```

Population density **per condition** in RNA pseudotime: 

```{r, fig.height=6, fig.width=8}
plts <- lapply(unique(colData(cds)$condition), function(x){
  whichsub <- colData(cds)$condition == x
  cds_sub <- cds[,whichsub]
  cds_sub@principal_graph_aux[["UMAP"]]$pseudotime <- cds_sub@principal_graph_aux[["UMAP"]]$pseudotime[whichsub]
  psuper(cds_sub, "cluster_name") + ggtitle(x)
})
patchwork::wrap_plots(plts)

```


But are the density of pseudotime different between the treatments and the control group ? We can test this with the two-sample Komogorov-Smirnov test, which tests the null hypothesis that the distribution of pseudotime for each cluster are drawn from the same distribution. 

```{r}
# none condition
cls <- levels(unique(colData(cds)$cluster_name))
whichsub <- colData(cds)$condition == "none"
cds_sub_none <- cds[,whichsub]
cds_sub_none@principal_graph_aux[["UMAP"]]$pseudotime <- cds_sub_none@principal_graph_aux[["UMAP"]]$pseudotime[whichsub]
pseudo_dt <- lapply(cls, function(y){
  data.frame("psuper" = cds_sub_none@principal_graph_aux[["UMAP"]]$pseudotime,
                        "label_input" = y)
})
pseudo_dt_none <- do.call("rbind", pseudo_dt)

kstst <- lapply(c("S1P", "LPS", "LPS_S1P"), function(x){
  
  # cds sub for each dondition
  whichsub <- colData(cds)$condition == x
  cds_sub <- cds[,whichsub]
  cds_sub@principal_graph_aux[["UMAP"]]$pseudotime <- cds_sub@principal_graph_aux[["UMAP"]]$pseudotime[whichsub]
  
  # preparing pseudotime in df
  pseudo_dt <- lapply(cls, function(y){
  data.frame("psuper" = cds_sub@principal_graph_aux[["UMAP"]]$pseudotime,
                          "label_input" = y)
  })
  pseudo_dt <- do.call("rbind", pseudo_dt)
  
  tsts <- lapply(cls, function(y){
    kstest <- ks.test(pseudo_dt$psuper[pseudo_dt$label_input == y], 
                      pseudo_dt_none$psuper[pseudo_dt_none$label_input == y])
    kstest$p.value
  })
  names(tsts) <- cls
  tsts <- unlist(tsts)
  tsts
})
names(kstst) <- c("S1P vs none", "LPS vs none", "LPS_S1P vs none")
kstst

```



```{r}
saveRDS(cds, "sce/monocle_rna.rds")

```



```{r}
sessionInfo()
```

